---
title: "FY20-23 CBP nationwide encounters"
author: "UWCHR"
date: "2024-01-31"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---

# CBP nationwide encounters: Data overview

https://www.cbp.gov/newsroom/stats/nationwide-encounters

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}

library(pacman)

p_load(here, tidyverse, zoo, lubridate, ggplot2, plotly, gghighlight)

cbp <- read_delim(here('analyze', 'input', 'cbp-nationwide-encounters-fy20-fy23-aor.csv'), delim=',')

names(cbp) %<>%
  str_replace_all(" \\s*\\([^\\)]+\\)", "") %>% 
  str_replace_all("\\s","_") %>%
  tolower()

month_abb_fy <- c("Oct", "Nov", "Dec", "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul","Aug", "Sep")

cbp <- cbp %>% 
  mutate(month_abb = ordered(str_to_title(month), month_abb_fy),
         month = match(month_abb, month.abb),
         calendar_year = case_when(month >= 10 ~ as.numeric(fiscal_year) -1,
                                   TRUE ~ as.numeric(fiscal_year)),
         fiscal_year = factor(fiscal_year))

cbp$year_mon <- as.yearmon(paste(cbp$calendar_year, cbp$month, sep='-'))
cbp$fy_quarter <- quarter(cbp$year_mon, fiscal_start=10, type="year.quarter")

skimr::skim(cbp)

```

# National data:

Note throughout "Encounters" may include multiple repeat encounters of the same individual/s.

CBP's portal presents data as year-over-year trend, we can replicate this presentation as below:

```{r national_total_fy, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}

p1 <- cbp %>% 
  group_by(fiscal_year, month_abb) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = month_abb, y = n, color = fiscal_year, group=fiscal_year)) + 
  geom_line() +
  ylim(0,NA) +
  labs(title = "CBP total national encounters, FY20-23")

p1

```

We can also visualize long-term trend:

```{r national_total, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}

p2 <- cbp %>% 
  group_by(year_mon) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = year_mon, y = n)) + 
  geom_line() +
  ylim(0,NA) +
  labs(title = "CBP total national encounters, FY20-23")

p2

```

CBP national encounters data are broken out by by `component`: U.S. Border Patrol (USBP, responsible for enforcement in interior and between ports of entry) or Office of Field Operations (OFO, a.k.a. customs, responsible for enforcement at ports of entry).

Components are broken out geographically by `aor` (area of responsibility), with USBP and OFO using mutually exclusive AOR categories with overlapping territories; e.g. USPB Spokane/Blaine sectors ("SPW", "BLW"), OFO Seattle/Portland Field Offices. (Data is also categorized by `land_border_region`).

```{r data_structure_aor, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}

# `aor` is a sub-set of `component`

cbp %>% 
  group_by(aor, component) %>% 
  summarize(n = sum(encounter_count))

```

Nationally, USBP reports greater proportion of encounters than OFO:

```{r national_breakdown_component, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}

p1 <- cbp %>% 
  group_by(component, fiscal_year, month_abb) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = month_abb, y = n, color = fiscal_year, group=fiscal_year)) + 
  geom_line() +
  facet_wrap(~component)

p1

```

Encounters are further broken out by `encounter_type` and `title_of_authority`: both OFO and USBP components report Title 42 "Expulsions"; while "USBP" reports "Apprehensions" and OFO reports "Inadmissibles" under Title 8 authority.

```{r data_structure_encounter_type, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}

# OFO encounters exclude apprehensions; USBP encounters exclude inadmissibles

cbp %>% 
  group_by(component, encounter_type) %>% 
  summarize(n = sum(encounter_count))

cbp %>% 
  group_by(component, title_of_authority) %>% 
  summarize(n = sum(encounter_count))

cbp %>% 
  group_by(encounter_type, title_of_authority) %>% 
  summarize(n = sum(encounter_count))

```

From Mar. 2020 through May 2023, both USBP and OFO carried out Title 42 expulsions in addition to Title 8 apprehensions/determinations of inadmissibility:

```{r national_breakdown_title, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}


# Note effect of Title 42 expulsions on rates of Title 8 apprehensions/inadmissibles

p2 <- cbp %>% 
  group_by(title_of_authority, fiscal_year, month_abb) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = month_abb, y = n, color = fiscal_year, group=fiscal_year)) + 
  geom_line() +
  facet_wrap(~title_of_authority)

p2

p2.1 <- cbp %>% 
  group_by(title_of_authority, year_mon) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = year_mon, y = n, color = title_of_authority, fill=title_of_authority)) + 
  geom_area() +
  ylim(0, NA) +
  labs(title = "Total CBP encounters by Title of Authority")

p2.1

```


```{r national_breakdown_region, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}

p3 <- cbp %>% 
  group_by(land_border_region, year_mon, month_abb) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = year_mon, y = n, color = land_border_region, group=land_border_region)) + 
  geom_line()

p3

```

We can also observe national trends with regards to CBP "Demographic" categories:

```{r national_breakdown_demog, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}

p3 <- cbp %>% 
  group_by(demographic, fiscal_year, month_abb) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = month_abb, y = n, color = fiscal_year, group=fiscal_year)) + 
  geom_line() +
  facet_wrap(~demographic)

p3

```

...and citizenship:

```{r cit, echo=FALSE, message=FALSE, include=TRUE}

cit <- cbp %>%
  group_by(citizenship) %>% 
  summarize(n = sum(encounter_count)) %>% 
  arrange(desc(n))

p1 <- cbp %>% 
  mutate(citizenship = case_when(
    citizenship %in% head(cit$citizenship, 15) ~ citizenship,
    TRUE ~ "ALL OTHERS"
  )) %>% 
  group_by(fiscal_year, citizenship) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x=fiscal_year, y=n, fill=citizenship)) +
  geom_col() +
  labs(title = "Total CBP encounters by country of citizenship (top 15)")

ggplotly(p1)

p2 <- cbp %>% 
  mutate(citizenship = case_when(
    citizenship %in% head(cit$citizenship, 15) ~ citizenship,
    TRUE ~ "ALL OTHERS"
  )) %>% 
  group_by(fiscal_year, title_of_authority, citizenship) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x=fiscal_year, y=n, fill=citizenship)) +
  geom_col() +
  labs(title = "Total CBP encounters by country of citizenship (top 15)") +
  facet_wrap(~title_of_authority)

ggplotly(p2)

```

```{r cit_rank, echo=FALSE, message=FALSE, include=TRUE}

cit_rank <- cbp %>% 
  group_by(fiscal_year, citizenship) %>% 
  summarize(n = sum(encounter_count)) %>% 
  arrange(fiscal_year, desc(n), citizenship) %>% 
  group_by(fiscal_year) %>% 
  mutate(ranking = row_number())

p1 <- cit_rank %>% 
  filter(ranking <= 10) %>% 
  ggplot(aes(x = fiscal_year, y = ranking, group = citizenship)) +
  geom_line(aes(color = citizenship, alpha = 1), size = 1) +
  geom_point(aes(color = citizenship, alpha = 1), size = 2) +
  scale_y_reverse() +
  labs(title = "Ranked country of citizenship for CBP encounters")

ggplotly(p1)

```

# Pacific Northwest USBP sectors/OFO field offices

Here, we focus on CBP components with jurisdiction in the Pacific Northwest: USBP Spokane and Blaine Sectors ("SPW", "BLW", respectively); and OFO Seattle and Portland Field Offices. Note OFO components report much greater proportion of encounters in PNW compared to USBP sectors.

```{r pnw_total, echo=FALSE, message=FALSE, include=TRUE}

p1 <- cbp %>% 
  filter(aor %in% c('BLW', 'SPW', "Seattle", "Portland")) %>% 
  group_by(component, year_mon, aor) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = year_mon, y = n, color = aor)) + 
  geom_line() +
  facet_wrap(~component) +
  labs(title="CBP encounters by component (OFO, USBP) and Field Office/Sector",
       caption="Source: CBP nationwide encounters FY2020-2023",
       color="Field Office/Sector")

p1


```

Northern border encounters up in general:
https://www.nbcnews.com/news/latino/more-asylum-requests-illegal-crossings-canada-northern-border-rcna126329

```{r blw_spw_total, echo=FALSE, message=FALSE, include=TRUE}

p1 <- cbp %>% 
  filter(aor %in% c('BLW', 'SPW')) %>% 
  group_by(aor, encounter_type, year_mon) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = year_mon, y = n, color = aor)) + 
  geom_line() +
  facet_wrap(~encounter_type)
 
p1

p2 <- cbp %>% 
  filter(aor %in% c('BLW', 'SPW')) %>% 
  group_by(aor,encounter_type, title_of_authority, year_mon) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = year_mon, y = n, color = aor)) + 
  geom_line() +
  facet_wrap(encounter_type~title_of_authority) +
  labs(title="USBP Blaine/Spokane Sector encounters",
       subtitle="Encounter type and Title of authority",
       caption="Source: CBP nationwide encounters FY2020-2023",
       color="Sector")

p2

```

Seattle OFO numbers parallel SEATAC passenger rates since 2020?
https://www.theurbanist.org/2023/04/13/airport-passenger-volumes-rebound-at-seatac-expansion-projects-roll-on/
https://www.seattletimes.com/life/travel/sea-tac-airport-employees-weathered-a-perfect-storm-this-winter/ 

```{r seattle_portland_total, echo=FALSE, message=FALSE, include=TRUE}

p1 <- cbp %>% 
  filter(aor %in% c('Seattle', 'Portland')) %>% 
  group_by(aor, encounter_type, year_mon) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = year_mon, y = n, color = aor)) + 
  geom_line() +
  facet_wrap(~encounter_type)
 
p1

p2 <- cbp %>% 
  filter(aor %in% c('Seattle', 'Portland')) %>% 
  group_by(aor,encounter_type, title_of_authority, year_mon) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = year_mon, y = n, color = aor)) + 
  geom_line() +
  facet_wrap(encounter_type~title_of_authority) +
  labs(title="OFO Seattle/Portland Field Office encounters",
       subtitle="Encounter type and Title of authority",
       caption="Source: CBP nationwide encounters FY2020-2023",
       color="Field Office")

p2

```

```{r seattle_portland_other, echo=FALSE, message=FALSE, include=TRUE}

p1 <- cbp %>% 
  filter(aor %in% c('Seattle', 'Portland')) %>% 
  group_by(fiscal_year, month_abb) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = month_abb, y = n, color = fiscal_year, group=fiscal_year)) + 
  geom_line()

p1

p2 <- cbp %>% 
  filter(aor %in% c('Seattle', 'Portland')) %>% 
  group_by(aor, fiscal_year, month_abb) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = month_abb, y = n, color = fiscal_year, group=fiscal_year)) + 
  geom_line() +
  facet_wrap(~aor)

p2

p3 <- cbp %>% 
  filter(aor %in% c('Seattle', 'Portland')) %>%
  group_by(aor, title_of_authority, fiscal_year, month_abb) %>% 
  summarize(n = sum(encounter_count)) %>% 
  ggplot(aes(x = month_abb, y = n, color = fiscal_year, group=fiscal_year)) + 
  geom_line() +
  facet_wrap(title_of_authority~aor)

p3
```

# Compare USBP reports

https://www.cbp.gov/document/stats/us-border-patrol-monthly-encounters-fy-2000-fy-2020

```{r usbp_monthly}

usbp <- read_delim(here('analyze', 'input', 'usbp_monthly_encounters_fy2000-fy2020.csv.gz'), delim='|')

usbp <- usbp %>% 
  dplyr::select(cy, fy, month, spokane, blaine) %>% 
  rename(usbp_spokane = spokane,
         usbp_blaine = blaine) %>% 
  mutate(fy = as.numeric(as.character(fy)))

cbp_spw_blw <- cbp %>% 
  rename(cy = calendar_year,
         fy = fiscal_year) %>% 
  filter(aor %in% c("SPW", "BLW")) %>% 
  group_by(fy, cy, month, aor) %>% 
  summarize(n = sum(encounter_count)) %>% 
  pivot_wider(names_from = aor, values_from = n) %>% 
  rename(cbp_spw = SPW,
         cbp_blw = BLW) %>% 
  mutate(fy = as.numeric(as.character(fy)))

dat <- full_join(usbp, cbp_spw_blw, by=c("cy", "fy", "month")) %>%
  mutate(diff_spw = usbp_spokane - cbp_spw,
         diff_blw = usbp_blaine - cbp_blw)

dat$blw <- coalesce(dat$usbp_blaine, dat$cbp_blw)
dat$spw <- coalesce(dat$usbp_spokane, dat$cbp_spw)

write_delim(dat, here('analyze', 'output', 'usbp_blw_spw_fy2000-fy2023.csv.gz'), delim='|')

stopifnot(sum(dat$diff_spw, na.rm=TRUE) == 0)
stopifnot(sum(dat$diff_blw, na.rm=TRUE) == 0)

dat_long <- dat %>% 
  dplyr::select(-starts_with(c('diff', 'cbp'))) %>% 
  rename('cbp_blw' = 'blw',
         'cbp_spw' = 'spw') %>% 
  pivot_longer(cols=-c("cy", "fy", "month")) %>% 
  mutate(year_mon = as.yearmon(paste(cy, month, sep='-')))

p1 <- dat_long %>% 
  ggplot(aes(x = year_mon, y = value, color = name)) +
  geom_line() +
  labs(title="USBP Blaine/Spokane Sector encounters FY2000-2023",
       caption="Source: USBP monthly encounters FY2000-2020\nCBP nationwide encounters FY2020-2023",
       color="Source/Sector")

p1

```

